Happiness Does Not Cause Altruism: A review of US General Social Survey data

### 2023-03-17 outputs/paper/paper.qmd
Added anchor links and structural fixes

---
title: "Happiness Does Not Cause Altruism: A review of US General Social Survey data"
format: pdf
author: "Sebastian Rodriguez, Laura Lee-Chu, Iz Leitch"
abstract: "The relationship between happiness and altruistic behaviors has been the subject of much research, with some studies suggesting a positive correlation and others finding no significant relationship. This paper examines the existing literature on the topic in addition to the survey results provided by NORC’s General Social Survey (GSS) of United States residents and analyzes the data to draw conclusions. Based on the available evidence, it appears that there is little to no relationship between happiness and altruistic behaviors, suggesting that individuals may engage in altruistic acts for reasons other than personal happiness." 
date: today
date-format: long
editor: visual
toc: TRUE
number-sections: TRUE
bibliography: references.bib
thanks: "Code and data supporting this analysis is available at: https://github.com/seb646/happiness-and-altruism"
---

@@

We wish to acknowledge the land on which this data was collected and analysed. For thousands of years, it has been the traditional land of the Huron-Wendat, the Seneca, and the Mississaugas of the Credit. Today, this meeting place is still the home to many Indigenous people from across Turtle Island and we are grateful to have the opportunity to work on this land.

@@

[^1]: General Happiness data is available at: https://gssdataexplorer.norc.org/variables/434/vshow

[^2]: Has Given Directions to a Stranger data is available at: https://gssdataexplorer.norc.org/variables/2886/vshow

[^3]: Has Given Money or Food to a Homeless Person data is available at: https://gssdataexplorer.norc.org/variables/2878/vshow

@@

#### Happiness
```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Raw GSS Data: General Happiness (by year)"
#| label: fig-5
#### Create a table summarizing happiness data from GSS ####
summarized_happiness_data <- merge(x = happy_positive_data, y = happy_negative_data, by = "year", all.x = TRUE) |>
   # Rename columns
  rename("Year" = "year", '"Very" or "Pretty" happy' = "people.x", '"Not too happy"' = "people.y")

@@

  # Increase width of first column
  column_spec(1, width = "10em")
```

#### Donating to the Homeless
```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Raw GSS Data: Giving Food or Money to the Homeless (by year)"
#| label: fig-6
#### Create a table summarizing homeless donation data from GSS ####
summarized_homeless_data <- merge(x = homeless_positive_data, y = homeless_negative_data, by = "year", all.x = TRUE) |>
   # Rename columns
  rename("Year" = "year", 'At least twice a year' = "people.x", 'Less than twice a year' = "people.y")

@@

  # Increase width of first column
  column_spec(1, width = "10em")
```

#### Giving Directions to a Stanger
```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Raw GSS Data: Giving Directions to a Stanger (by year)"
#| label: fig-7
#### Create a table summarizing giving strangers directions data from GSS ####
summarized_directions_data <- merge(x = directions_positive_data, y = directions_negative_data, by = "year", all.x = TRUE) |>
   # Rename columns
  rename("Year" = "year", 'At least twice a year' = "people.x", 'Less than twice a year' = "people.y")

@@

  # Increase width of first column
  column_spec(1, width = "10em")
```

### Appendix B: Supplementary Survey
The supplementary survey attached to this project is available through Google Forms: https://forms.gle/Zr2jj93bXFaaPkPdA.

#### Survey Introduction 
Researchers have long sought to find a connection between happiness and altruistic actions. Some studies show that happiness may lead to greater levels of altruism, while others fail to find a relationship between the emotion and subsequent action. With homelessness on the rise and a growing separation between socio-economic classes, it is important to discover the leading causes of altruism.

@@

If you have any questions or concerns, feel free to contact Sebastian Rodriguez (s [dot] rodriguez [at] mail [dot] utoronto [dot] ca).

#### Survey Questions
1.  On a scale of 1-5, how happy do you feel right now?
    a. Very happy
    b. Pretty happy
    c. Happy
    d. Not too happy
    e. Unhappy
2.  How often have you given directions to a stranger in the past year?
    a. More than once a week
    b. Once a week
    c. Once a month 
    d. At least 2 or 3 times in the past year
    e. Once in the past year
    f. Not at all in the past year
3.  How often are you given the opportunity to give directions to a stranger?
    a. More than once a week
    b. Once a week
    c. Once a month 
    d. At least 2 or 3 times in the past year
    e. Once in the past year
    f. Not at all in the past year
4.  If you did not give directions to a stranger, did you want to?
    a. Yes
    b. It depends on the day
    c. It depends on the time I have in the moment
    d. No
5.  If you had the desire but did not give directions to a stranger, what prevented you from doing so?
    a. I do not live in an area with a lot of strangers asking for directions
    b. I did not have the time to help
    c. I did not know enough about the area to be helpful
    d. Another factor beyond my control prevented me from helping
6.  How often have you donated food or money to a homeless person in the past year?
    a. More than once a week
    b. Once a week
    c. Once a month 
    d. At least 2 or 3 times in the past year
    e. Once in the past year
    f. Not at all in the past year
7.  How often are you given the opportunity to donate to a homeless person?
    a. More than once a week
    b. Once a week
    c. Once a month 
    d. At least 2 or 3 times in the past year
    e. Once in the past year
    f. Not at all in the past year
8.  If you did not donate to a homeless person, did you want to?
    a. Yes
    b. It depends on the day
    c. It depends on the money and resources at my disposal
    d. No
9.  If you had the desire but did not donate to a homeless person, what prevented you from doing so?
    a. I do not live in an area with a lot of homeless people
    b. I did not have the time to help
    c. I did not have the money or resources to help
    d. Another factor beyond my control prevented me from helping

#### Survey Conclusion
Thank you for participating in this survey! Your responses will greatly aid our effort to understand altruism and whether happiness has any impact on the behavior. If at any time you wish to withdraw from this survey, please contact Sebastian Rodriguez (s [dot] rodriguez [at] mail [dot] utoronto [dot] ca).

>>>>

---
title: "Happiness Does Not Cause Altruism: A review of US General Social Survey data"
format: pdf
author: "Sebastian Rodriguez, Laura Lee-Chu, Iz Leitch"
abstract: "The relationship between happiness and altruistic behaviors has been the subject of much research, with some studies suggesting a positive correlation and others finding no significant relationship. This paper examines the existing literature on the topic in addition to the survey results provided by NORC’s General Social Survey (GSS) of United States residents and analyzes the data to draw conclusions. Based on the available evidence, it appears that there is little to no relationship between happiness and altruistic behaviors, suggesting that individuals may engage in altruistic acts for reasons other than personal happiness." 
date: today
date-format: long
editor: visual
toc: TRUE
number-sections: TRUE
bibliography: references.bib
thanks: "Code and data supporting this analysis is available at: [https://github.com/seb646/happiness-and-altruism](https://github.com/seb646/happiness-and-altruism)"
---

@@

We wish to acknowledge the land on which this data was analysed. For thousands of years, it has been the traditional land of the Huron-Wendat, the Seneca, and the Mississaugas of the Credit. Today, this meeting place is still the home to many Indigenous people from across Turtle Island and we are grateful to have the opportunity to work on this land.

@@

[^1]: General Happiness data is available at: [https://gssdataexplorer.norc.org/variables/434/vshow](https://gssdataexplorer.norc.org/variables/434/vshow)

[^2]: Has Given Directions to a Stranger data is available at: [https://gssdataexplorer.norc.org/variables/2886/vshow](https://gssdataexplorer.norc.org/variables/2886/vshow)

[^3]: Has Given Money or Food to a Homeless Person data is available at: [https://gssdataexplorer.norc.org/variables/2878/vshow](https://gssdataexplorer.norc.org/variables/2878/vshow)\

@@

**Happiness**

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Raw GSS Data: General Happiness (by year)"
#| label: fig-5
#### Create a table summarizing happiness data from GSS ####
summarized_happiness_data <- merge(x = happy_positive_data, y = happy_negative_data, by = "year", all.x = TRUE) |>
   # Rename columns
  rename("Year" = "year", '"Very" or "Pretty" happy' = "people.x", '"Not too happy"' = "people.y")

@@

  # Increase width of first column
  column_spec(1, width = "10em")
```
\

**Donating to the Homeless**
```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Raw GSS Data: Giving Food or Money to the Homeless (by year)"
#| label: fig-6
#### Create a table summarizing homeless donation data from GSS ####
summarized_homeless_data <- merge(x = homeless_positive_data, y = homeless_negative_data, by = "year", all.x = TRUE) |>
   # Rename columns
  rename("Year" = "year", 'At least twice a year' = "people.x", 'Less than twice a year' = "people.y")

@@

  # Increase width of first column
  column_spec(1, width = "10em")
```
\

**Giving Directions to a Stranger**

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Raw GSS Data: Giving Directions to a Stanger (by year)"
#| label: fig-7
#### Create a table summarizing giving strangers directions data from GSS ####
summarized_directions_data <- merge(x = directions_positive_data, y = directions_negative_data, by = "year", all.x = TRUE) |>
   # Rename columns
  rename("Year" = "year", 'At least twice a year' = "people.x", 'Less than twice a year' = "people.y")

@@

  # Increase width of first column
  column_spec(1, width = "10em")
```
\newpage

### Appendix B: Supplementary Survey
The supplementary survey attached to this project is available through Google Forms: [https://forms.gle/Zr2jj93bXFaaPkPdA](https://forms.gle/Zr2jj93bXFaaPkPdA). \

**Survey Introduction**

Researchers have long sought to find a connection between happiness and altruistic actions. Some studies show that happiness may lead to greater levels of altruism, while others fail to find a relationship between the emotion and subsequent action. With homelessness on the rise and a growing separation between socio-economic classes, it is important to discover the leading causes of altruism.

@@

If you have any questions or concerns, feel free to contact Sebastian Rodriguez (s [dot] rodriguez [at] mail [dot] utoronto [dot] ca). \

**Survey Questions**

1.  On a scale of 1-5, how happy do you feel right now?
    a. Very happy
    b. Pretty happy
    c. Happy
    d. Not too happy
    e. Unhappy
2.  How often have you given directions to a stranger in the past year?
    a. More than once a week
    b. Once a week
    c. Once a month 
    d. At least 2 or 3 times in the past year
    e. Once in the past year
    f. Not at all in the past year
3.  How often are you given the opportunity to give directions to a stranger?
    a. More than once a week
    b. Once a week
    c. Once a month 
    d. At least 2 or 3 times in the past year
    e. Once in the past year
    f. Not at all in the past year
4.  If you did not give directions to a stranger, did you want to?
    a. Yes
    b. It depends on the day
    c. It depends on the time I have in the moment
    d. No
5.  If you had the desire but did not give directions to a stranger, what prevented you from doing so?
    a. I do not live in an area with a lot of strangers asking for directions
    b. I did not have the time to help
    c. I did not know enough about the area to be helpful
    d. Another factor beyond my control prevented me from helping
6.  How often have you donated food or money to a homeless person in the past year?
    a. More than once a week
    b. Once a week
    c. Once a month 
    d. At least 2 or 3 times in the past year
    e. Once in the past year
    f. Not at all in the past year
7.  How often are you given the opportunity to donate to a homeless person?
    a. More than once a week
    b. Once a week
    c. Once a month 
    d. At least 2 or 3 times in the past year
    e. Once in the past year
    f. Not at all in the past year
8.  If you did not donate to a homeless person, did you want to?
    a. Yes
    b. It depends on the day
    c. It depends on the money and resources at my disposal
    d. No
9.  If you had the desire but did not donate to a homeless person, what prevented you from doing so?
    a. I do not live in an area with a lot of homeless people
    b. I did not have the time to help
    c. I did not have the money or resources to help
    d. Another factor beyond my control prevented me from helping\

**Survey Conclusion**

Thank you for participating in this survey! Your responses will greatly aid our effort to understand altruism and whether happiness has any impact on the behavior. If at any time you wish to withdraw from this survey, please contact Sebastian Rodriguez (s [dot] rodriguez [at] mail [dot] utoronto [dot] ca).
